---
title: "Custom analyses LY mAbs"
author: "Tyler Starr"
date: "2/6/2021"
output:
  github_document:
    toc: true
    html_preview: false
editor_options: 
  chunk_output_type: inline
---

This notebook does some random analyses on the clinical antibodies set that vary from the more constrained global pipeline.

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("yaml","data.table","tidyverse","ggrepel","bio3d","gridExtra")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("./config.yaml")

#read in escape profiles file
profiles_config <- read_yaml(file=config$escape_profiles_config)

#make output directory
output_dir <- config$custom_plots_dir
if(!file.exists(output_dir)){
  dir.create(file.path(output_dir))
}
```
Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

## Data input and formatting

Read in escape fractions, rename some things to make them easier to work with.

```{r data input}
scores <- data.table(read.csv(file=config$escape_fracs,stringsAsFactors=F))

scores <- scores[selection %in% names(profiles_config$`LY_cocktail`$conditions) & library=="average", .(selection,condition,site,protein_site,wildtype,mutation,mut_escape_frac_epistasis_model,site_total_escape_frac_epistasis_model)]

setnames(scores,"mut_escape_frac_epistasis_model","mut_escape_frac");setnames(scores,"site_total_escape_frac_epistasis_model","site_total_escape")

scores[,antibody:=as.character(profiles_config$`LY_cocktail`$conditions[condition])]

scores <- scores[,.(antibody,site,protein_site,wildtype,mutation,mut_escape_frac,site_total_escape)]

scores[,site_max_escape:=max(mut_escape_frac,na.rm=T),by=c("antibody","site")]


```

## Part 1: Circulating variants at the per-mut level

Current notebook on circulating mutations considers all mutations at a site together, regardless of the escape conferred by the particular mutation that is circulating. Make plots at the per-mut level.

We read in table reporting circulating variants. We add new columns to our data frame indicating the nobs and frequency on GISAID, and the number of countries in which a mutant has been observed. Then, for each antibody, we plot per-mutation escape fraction versus frequency (log10), with a 'pseudo-frequency' of 0.1x the lowest actual frequency, to enable log10 plotting)

```{r mutation_escape_v_freq, echo=T, fig.width=10, fig.height=3.75, fig.align="center", dpi=300,dev="png"}
#read in table giving mutant frequencies on GISAID
counts <- read.csv(config$gisaid_mutation_counts,stringsAsFactors=F)
#add to scores table
scores[,count:=0];scores[,n_countries:=0];scores[,frequency:=0]
for(i in 1:nrow(counts)){
  scores[protein_site==counts[i,"site"] & mutation==counts[i,"mutant"],count:=counts[i,"count"]]
  scores[protein_site==counts[i,"site"] & mutation==counts[i,"mutant"],n_countries:=counts[i,"n_countries"]]
  scores[protein_site==counts[i,"site"] & mutation==counts[i,"mutant"],frequency:=counts[i,"frequency"]]
}

#set factor order for antibodies
scores$antibody <- factor(scores$antibody,levels=c("LY-CoV555","LY-CoV016","LY-CoV555+LY-CoV016"))

scores[,pseudo_frequency:=frequency]
scores[frequency==0,pseudo_frequency:=0.1*min(scores[frequency>0,frequency])]

p1 <- ggplot(scores)+aes(x=pseudo_frequency,y=mut_escape_frac)+
  geom_point(shape=16, alpha=0.5, size=2.25)+
  facet_wrap(~antibody,nrow=1)+
  scale_x_continuous(trans="log10")+
  scale_y_continuous(limits=c(0,1.05))+
  theme_classic()+
  xlab('mutant frequency on GISAID (log10 scale)')+
  ylab('mutant escape fraction')+
  geom_text_repel(aes(label=ifelse((mut_escape_frac>0.15 & frequency>1e-6) | (mut_escape_frac>0.05 & frequency>1e-3),as.character(paste(wildtype,protein_site,mutation,sep="")),'')),size=3,color="gray40")
p1

invisible(dev.print(pdf, paste(output_dir,"/circ-mut-scatter_mAbs_LY.pdf",sep="")))

```

## Part 2: Distribution of functional effects of mutations circulating at different levels

For contextualizing ACE2 and expression deficits of antigenic mutations, what is the distribution of these phenotypes among sequences on GISAID?

```{r GISAID_DFE, echo=T, fig.width=7, fig.height=4, fig.align="center", dpi=300,dev="png"}
#read in DMS scores
dms <- data.table(read.csv(file=config$mut_bind_expr,stringsAsFactors = F))
setnames(dms,"site_SARS2","site")
dms <- dms[mutant != "*",]
dms <- dms[as.character(wildtype) != as.character(mutant),]

dms[,count:=0]
for(i in 1:nrow(counts)){
  dms[site==counts[i,"site"] & mutant==counts[i,"mutant"],count:=counts[i,"count"]]
}

#define factor for different nobs cutoffs, and collate into long data table for plotting
muts_temp <- dms
muts_temp$nobs_indicator<-factor("all muts",levels=c("all muts",">=1", ">=10",">=25"))
muts_temp_add0 <- dms[count>=1, ]
muts_temp_add0$nobs_indicator<-factor(">=1",levels=c("all muts",">=1", ">=10",">=25"))
muts_temp_add10 <- dms[count>=10, ]
muts_temp_add10$nobs_indicator<-factor(">=10",levels=c("all muts",">=1", ">=10",">=25"))
muts_temp_add25 <- dms[count>=25, ]
muts_temp_add25$nobs_indicator<-factor(">=25",levels=c("all muts",">=1", ">=10",">=25"))

muts_temp <- rbind(muts_temp,muts_temp_add0,muts_temp_add10,muts_temp_add25)
muts_temp$nobs_indicator <- factor(muts_temp$nobs_indicator, levels=c("all muts",">=1", ">=10",">=25"))

set.seed(198)
p1 <- ggplot(muts_temp[!is.na(muts_temp$bind_avg),],aes(x=nobs_indicator,y=bind_avg))+
  geom_boxplot(outlier.shape=16, width=0.4, outlier.alpha=0.5)+
  #geom_jitter(width=0.2, alpha=0.1, shape=16)+
  xlab("mutation count in GISAID sequences")+ylab("mutation effect on binding")+
  theme_classic()+geom_hline(yintercept=-2.35, linetype = 'dotted', col = 'gray50')+
  geom_hline(yintercept=0,linetype='dashed',col='gray50')

p2 <- ggplot(muts_temp[!is.na(muts_temp$expr_avg),],aes(x=nobs_indicator,y=expr_avg))+
  geom_boxplot(outlier.shape=16, width=0.4, outlier.alpha=0.5)+
  #geom_jitter(width=0.2, alpha=0.1, shape=16)+
  xlab("mutation count in GISAID sequences")+ylab("mutation effect on expression")+
  theme_classic()+geom_hline(yintercept=-1, linetype = 'dotted', col = 'gray50')+
  geom_hline(yintercept=0,linetype='dashed',col='gray50')

grid.arrange(p1,p2,ncol=2)

invisible(dev.print(pdf, paste(output_dir,"/distribution-ACE2-expr-dms_v_nobs-GISAID.pdf",sep="")))

```


## Part 3: escape within or without the structural epitope

We want to try a few plots -- at the per-mut and per-site level, compare distribution of escape scores between structural contacts and non-contacts. We should try the contact/non-contact, but also a "close" category corresponding to 4-8A distance? Means we'll probably want to re-do the structural analysis here within this "custom analyses" script.

First, load pdbs, and generate list of residues that are <4A or 4-8A from each mAb

```{r define_structural_contacts}
pdb_LY555 <- read.pdb(file="./data/pdbs/7kmg_single.pdb")
pdb_LY016 <- read.pdb(file="./data/pdbs/7c01_single.pdb")

contacts_LY555_4A <- binding.site(pdb_LY555, 
                                      a.inds=atom.select(pdb_LY555, chain="C"),
                                      b.inds=atom.select(pdb_LY555,chain=c("A","B")),
                                      cutoff=4,hydrogens=F)$resno
contacts_LY555_4to8A <- binding.site(pdb_LY555, 
                                      a.inds=atom.select(pdb_LY555, chain="C"),
                                      b.inds=atom.select(pdb_LY555,chain=c("B","D")),
                                      cutoff=8,hydrogens=F)$resno
contacts_LY555_4to8A <- contacts_LY555_4to8A[!(contacts_LY555_4to8A %in% contacts_LY555_4A)]


contacts_LY016_4A <- binding.site(pdb_LY016, 
                                      a.inds=atom.select(pdb_LY016, chain="A"),
                                      b.inds=atom.select(pdb_LY016,chain=c("H","L")),
                                      cutoff=4,hydrogens=F)$resno
contacts_LY016_4to8A <- binding.site(pdb_LY016, 
                                      a.inds=atom.select(pdb_LY016, chain="A"),
                                      b.inds=atom.select(pdb_LY016,chain=c("H","L")),
                                      cutoff=8,hydrogens=F)$resno
contacts_LY016_4to8A <- contacts_LY016_4to8A[!(contacts_LY016_4to8A %in% contacts_LY016_4A)]

```
Next, make jitterplots showing site total escape for the three categories of structural class.

```{r plot_total_escape_contacts, echo=T, fig.width=4, fig.height=3, fig.align="center", dpi=300,dev="png"}
temp <- scores[antibody %in% c("LY-CoV016","LY-CoV555"),]
#add column whether a site is in direct contact, proximal, or distal to interface
temp[,contact:="distal"]
temp[antibody=="LY-CoV016" & protein_site %in% contacts_LY016_4A,contact:="contact"]
temp[antibody=="LY-CoV016" & protein_site %in% contacts_LY016_4to8A,contact:="proximal"]
temp[antibody=="LY-CoV555" & protein_site %in% contacts_LY555_4A,contact:="contact"]
temp[antibody=="LY-CoV555" & protein_site %in% contacts_LY555_4to8A,contact:="proximal"]
temp$contact <- factor(temp$contact,levels=c("contact","proximal","distal"))

#add value indicating the heuristic cutoff for "strong" escape for each mAb
cutoff_LYCoV016 <- max(5*median(temp[antibody=="LY-CoV016",site_total_escape]),0.05*max(temp[antibody=="LY-CoV016",site_total_escape]))
cutoff_LYCoV555 <- max(5*median(temp[antibody=="LY-CoV555",site_total_escape]),0.05*max(temp[antibody=="LY-CoV555",site_total_escape]))

temp[antibody=="LY-CoV016",cutoff:=cutoff_LYCoV016]
temp[antibody=="LY-CoV555",cutoff:=cutoff_LYCoV555]

#sitewise
ggplot(unique(temp[,.(antibody,protein_site,site_total_escape,contact)]),aes(x=contact,y=site_total_escape))+
  geom_jitter(width=0.15, alpha=0.5,height=0,size=2.25,shape=16)+
  geom_hline(data=unique(temp[,.(antibody,cutoff)]),aes(yintercept=cutoff),linetype="dashed",color="gray50")+
  theme_classic()+
  theme(axis.text.x=element_text(angle=90,hjust=1))+
  xlab("structural context")+ylab("sitewise total escape")+
  facet_wrap(~antibody,nrow=1)

invisible(dev.print(pdf, paste(output_dir,"/escape_by_contact_type.pdf",sep="")))

```
Table with number of sites in each structural category, and number of sites of "strong" escape

```{r table_strong_sites_by_structure}
temp <- unique(temp[,.(antibody,protein_site,site_total_escape,contact,cutoff)])

dt <- data.frame(antibody=c("LY-CoV555","LY-CoV016"),n_contact=NA,n_contact_escape=NA,n_proximal=NA,n_proximal_escape=NA,n_distal=NA,n_distal_escape=NA)
for(i in 1:nrow(dt)){
  dt$n_contact[i] <- nrow(temp[as.character(antibody)==dt[i,"antibody"] & contact=="contact",])
  dt$n_contact_escape[i] <- nrow(temp[as.character(antibody)==dt[i,"antibody"] & contact=="contact" & site_total_escape>cutoff,])
  dt$n_proximal[i] <- nrow(temp[as.character(antibody)==dt[i,"antibody"] & contact=="proximal",])
  dt$n_proximal_escape[i] <- nrow(temp[as.character(antibody)==dt[i,"antibody"] & contact=="proximal" & site_total_escape>cutoff,])
  dt$n_distal[i] <- nrow(temp[as.character(antibody)==dt[i,"antibody"] & contact=="distal",])
  dt$n_distal_escape[i] <- nrow(temp[as.character(antibody)==dt[i,"antibody"] & contact=="distal" & site_total_escape>cutoff,])
}
kable(dt)
```

